\large How resource abundance and stochasticity affect organisms' range sizes

### 2024-07-31 writing/manuscript.Rmd
formatted for ME, added abbreviations, fixed typos

```{r, echo=FALSE}
abstract <- "The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. Since organisms adjust to variable environmental conditions, failing to consider the effects of resource (un)predictability can result in an insufficient understanding of an organism's range size, which results in challenges for both ecological theory and applied conservation. In this study, we leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on range sizes. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resources interact to shape predictable patterns in range size. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
```

@@

\noindent \textbf{Key words:} energetics, resource abundance, environmental stochasticity, home range, range size, `ctmm`

<!-- Authors' emails:
Stefano Mezzini: stefano.mezzini@ubc.ca
Chris H. Fleming: christen.fleming@ucf.edu
E. Patrícia Medici: medici@ipe.org.br
Michael J. Noonan: michael.noonan@ubc.ca
-->

@@

# Introduction

@@

While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976; @nathan_movement_2008]. The relationship between organisms' movement and resource abundance has long been of interest to biologists. In his seminal paper, @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis. Many researchers have since demonstrated that organisms adapt their range sizes in response to resources abundance, but results are typically reported as independent, novel findings. Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021; @broekman_environmental_2024]. Thus, there remains a need for clear hypotheses of the effects of both resource abundance and stochasticity on organisms’ range sizes.

@@

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses, which is similar to that presented by @harestad_home_1979. When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. Like @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$. These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, switches from range residency to nomadism (or vice-versa) will occur over evolutionary timescales rather than over an organism's lifespan (Fig. \@ref(fig:mu-hyp)), since larger ranges require greater vagility, which, in turn, is facilitated by the development of morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

Each of the 200 tracks were placed on a grid with common starting point $\langle 0, 0\rangle$ (fig. B1). Each time the simulated individual moved to a new cell, it collected $R$ resources sampled from a Gamma distribution. The mean and variance of the distribution were defined by a series of deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig. \@ref(fig:5-5-sims)). The value of $t$ was constant within each set of 200 tracks, so the distribution $R$ was sampled from was independent of both the organism's location and its time spent moving. Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track (section 2.1 of Appendix B). Finally, we fit an OUF movement model [@fleming_fine-scale_2014] to the set of tracks to calculate the 95% Gaussian home-range size using the formula $$H_{95\%} = -2 \log(1 - 0.95) \pi \hat \varsigma^2,$$ where $\hat \varsigma^2$ is the positional variance estimated by the movement model.

@@

```{r 5-5-sims, fig.cap="Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogenous over space for a given $t$). Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases. Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

@@

\noindent The simulations in the section above support the hypotheses we presented in the introduction, but they are based on assumptions that are often not met in real natural environments. Organisms live in spatiotemporally heterogeneous and dynamic environments that promote the use of perceptual ranges, navigation, and memory. Together, these abilities result in selective space use that depends on resource availability [@kacelnik_ideal_1992] and resource depletion [@charnov_optimal_1976].

@@

Fig. \@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ [telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022]. Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period. The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, @wood_smoothing_2016] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475). Panel C shows the changes in the tapir's 7-day home range over time. Note how the tapir uses more space during periods of lower NDVI (e.g., August 2017) and less space during periods with high NDVI (January 2018). Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018). Finally, panels D and E show the estimated (marginal) effects of $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ on the tapir's 7-day home-range size. Since $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of $R$ on $H$ should be modeled carefully. To avoid over-fitting the model, we constrained the smooth effects of $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ and their interaction effect to a small basis size ($\tt{k = 3}$). Additional information is provided in appendix C. The results presented in panels D-F of fig. \@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\mu(t, \vec u)$ and increases with $\sigma^2(t, \vec u)$, and the effect of $\mu(t, \vec u)$ depends on $\sigma^2(t, \vec u)$, and vice-versa. Alone, $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F). Additionally, note how high NDVI values (> 0.8) cause $\sigma^2(t, \vec u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.

```{r tapir-mw, fig.cap="Effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on the home-range size of a lowland tapir (\\emph{Tapirus terrestris}). (A) Trends in resource abundance over time, $\\mu(t, \\vec u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period. (B) Varince in resources over time, $\\sigma^2(t, \\vec u)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period. (C) Seven-day 95% home range estimated using Autocorrelated Kernel Density Estimation. (D, E) Estimated marginal effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on home-range size. The model accounted for the marginal effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ and their interaction effect. (F) Estimated home-range size in response to changes in both $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$. Note how the effect of $\\sigma^2(t, \\vec u)$ is more pronounced when $\\mu(t, \\vec u)$ is low. See Appendix C for additional information. The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici \\emph{et al.} (2022).", out.width='100%'}
knitr::include_graphics('../figures/tapir-example.png')
```

@@

Deviations from the simulations offer a means of detecting when the underlying assumptions are inappropriate and how additional factors may affect organisms' responses to changes in $\e(R)$ and $\var(R)$. For example, energetic costs of movement are often non-negligible and depend on organism size [@taylor_energetics_1982], movement speed [@taylor_energetics_1982], and ambient temperature [@brown_toward_2004; @hou_cold_2020]. In addition, an organism may alter its movement behavior, physiology, and energetic needs to buffer itself against changes in $\e(R)$ and $\var(R)$ by using space selectively [@johnson_comparison_1980; @rickbeil_plasticity_2019; @riotte-lambert_environmental_2020; @ranc_memory_2022] and adapting their behavior and physiology over time [@southwood_habitat_1977; @levins_evolution_1974]. Before or during periods of scarcity, organisms may cache resources [@nespolo_why_2022], build up fat reserves [@middleton_green-wave_2018], enter states of dormancy [@goldberg_hibernation_2021; @reher_short_2018; @mohr_cellular_2020], or even pause fetal growth [@schmidt_interplay_2020]. However, organisms may be unable to respond to changes in $\e(R)$ and $\var(R)$ optimally due to various reasons, including limited perceptive range [@steixner-kumar_strategies_2020], lack of experience [@foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @merkle_spatial_2019; @fagan_spatial_2013; @abrahms_memory_2019], avoidance of competitors and predators [@kacelnik_ideal_1992; @fretwell_territorial_1969], or a physiology that is not amenable to things like hibernation or fat storage. Thus organisms may relocate their range to a sub-optimal location [@ciuti_effects_2012; @rich_anthropogenic_2012; @burson_competition_2018; @torrez-herrera_monkeys_2020], which may exacerbate the effects of $\e(R)$ and $\var(R)$ on both mean range size and the variance around it.

@@

# Conclusion

@@

# Acknowledgements

\noindent We would like to thank Dr. Simon Wood for providing code to fit a Beta location-scale GAM despite not being involved directly with the project. Additionally, we thank all those who provided feedback on all posters, presentation, and writings related to this project. In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely, in alphabetical order by first name: Aimee Chhen, Jessa Marley, Kim Hinz, Lauren Mills, Sarah Wyse, and Dr. Simon Wood. This work was supported by the NSERC Discovery Grant RGPIN-2021-02758 to MJN and funds from the University of British Columbia Okanagan, the Canadian Foundation for Innovation, Mitacs, and BC Parks.

# Code and data availability

All code and data used for this manuscript is available on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch](https://github.com/QuantitativeEcologyLab/hr-resource-stoch), with the exception of the tapir data, which is available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).

# Conflict of interest

\noindent The authors declare there are no conflicts of interest.

>>>>

```{r, echo=FALSE}
abstract <- "\\textbf{Background:} The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. Since organisms adjust to variable environmental conditions, failing to consider the effects of resource (un)predictability can result in an insufficient understanding of an organism's range size, which results in challenges for both ecological theory and applied conservation. \\textbf{Methods:} We leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on range sizes. We then use simulated movement data using continuous-time movement models to demonstrate how the combined effects of mean and variance in resources interact to shape predictable patterns in range size. Finally, we test the two hypotheses using real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado. \\textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity. Additionally, the effects of resource stochasticity depend strongly on resource abundance, and organisms' responses to resource abundance and stochasticity can be complex, since the distribution and predictability of resources can exacerbate the effects of other drivers of movement, such as resource depletion, competition, and predation. \\textbf{Conclusions:} Accounting for resource abundance and stochasticity is crucial for understanding the movement behavior of free-ranging organisms. Failing to account for resource stochasticity can lead to an incomplete and incorrect understanding of how and why organisms move, particularly during periods of rapid change."
# stringi::stri_count_words(abstract)
```

@@

\noindent \textbf{Key words:} energetics, resource abundance, resource stochasticity, environmental stochasticity, home range, range size, movement behavior, `ctmm`

@@

# Background

@@

While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976; @nathan_movement_2008]. The relationship between organisms' movement and resource abundance has long been of interest to biologists. In his seminal paper, Burt @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, Southwood @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce. Two years later, Harestad and Bunnel @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption rate (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis. Many researchers have since demonstrated that organisms adapt their range sizes in response to resources abundance, but results are typically reported as independent, novel findings. Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021; @broekman_environmental_2024]. Thus, there remains a need for clear hypotheses of the effects of both resource abundance and stochasticity on organisms’ range sizes.

@@

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses, which is similar to that presented by @harestad_home_1979. When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. Like Harestad and Bunnel @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$. These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, switches from range residency to nomadism (or vice-versa) will occur over evolutionary timescales rather than over an organism's lifespan (Fig. \@ref(fig:mu-hyp)), since larger ranges require greater vagility, which, in turn, is facilitated by the development of morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

Each of the 200 tracks were placed on a grid with common starting point $\langle 0, 0\rangle$ (fig. B1). Each time the simulated individual moved to a new cell, it collected $R$ resources sampled from a Gamma distribution. The mean and variance of the distribution were defined by a series of deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig. \@ref(fig:5-5-sims)). The value of $t$ was constant within each set of 200 tracks, so the distribution $R$ was sampled from was independent of both the organism's location and its time spent moving. Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track (section 2.1 of Appendix B). Finally, we fit an OUF movement model [@fleming_fine-scale_2014] to the set of tracks to calculate the 95% Gaussian home-range size using the formula $$\hat{H}_{95\%} = -2 \log(1 - 0.95) \pi \hat \varsigma^2,$$ where $\hat \varsigma^2$ is the positional variance estimated by the movement model.

@@

```{r 5-5-sims, fig.cap="Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogeneous over space for a given $t$). Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases. Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

@@

\noindent The simulations in the section above support the hypotheses we presented in the background section, but they are based on assumptions that are often not met in real natural environments. Organisms live in spatiotemporally heterogeneous and dynamic environments that promote the use of perceptual ranges, navigation, and memory. Together, these abilities result in selective space use that depends on resource availability [@kacelnik_ideal_1992] and resource depletion [@charnov_optimal_1976].

@@

Fig. \@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in estimated $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ (telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022). Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period. The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, @wood_smoothing_2016] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475). Panel C shows the changes in the tapir's 7-day home range over time. Note how the tapir uses more space during periods of lower NDVI (e.g., August 2017) and less space during periods with high NDVI (January 2018). Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018). Finally, panels D and E show the estimated (marginal) effects of $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ on the tapir's 7-day home-range size. Since $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of $R$ on $H$ should be modeled carefully. To avoid over-fitting the model, we constrained the smooth effects of $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ and their interaction effect to a small basis size ($\tt{k = 3}$). Additional information is provided in appendix C. The results presented in panels D-F of fig. \@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\hat\mu(t, \vec u)$ and increases with $\hat\sigma^2(t, \vec u)$, and the effect of $\hat\mu(t, \vec u)$ depends on $\hat\sigma^2(t, \vec u)$, and vice-versa. Alone, $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F). Additionally, note how high NDVI values ($\hat \mu(t, \vec u) > 0.8$) cause $\hat\sigma^2(t, \vec u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.

```{r tapir-mw, fig.cap="Effects of estimated $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on the home-range size of a lowland tapir (\\emph{Tapirus terrestris}). (A) Trends in resource abundance over time, $\\hat\\mu(t, \\vec u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period. (B) Variance in resources over time, $\\hat\\sigma^2(t, \\vec u)$, estimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period. (C) Seven-day 95% home range estimated using Autocorrelated Kernel Density Estimation. (D, E) Estimated marginal effects of $\\hat\\mu(t, \\vec u)$ and $\\hat\\sigma^2(t, \\vec u)$ on home-range size. The model accounted for the marginal effects of $\\hat\\mu(t, \\vec u)$, $\\hat\\sigma^2(t, \\vec u)$, and their interaction effect. (F) Estimated home-range size in response to changes in both $\\hat\\mu(t, \\vec u)$ and $\\hat\\sigma^2(t, \\vec u)$. Note how the effect of $\\hat\\sigma^2(t, \\vec u)$ is more pronounced when $\\hat\\mu(t, \\vec u)$ is low. See Appendix C for additional information. The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici \\emph{et al.} [27].", out.width='100%'}
knitr::include_graphics('../figures/tapir-example.png')
```

@@

Deviations from the simulations offer a means of detecting when the underlying assumptions are inappropriate and how additional factors may affect organisms' responses to changes in $\e(R)$ and $\var(R)$. For example, energetic costs of movement are often non-negligible and depend on organism size [@taylor_energetics_1982], movement speed [@taylor_energetics_1982], and ambient temperature [@brown_toward_2004; @hou_cold_2020]. In addition, an organism may alter its movement behavior, physiology, and energetic needs to buffer itself against changes in $\e(R)$ and $\var(R)$ by using space selectively [@johnson_comparison_1980; @rickbeil_plasticity_2019; @riotte-lambert_environmental_2020; @ranc_memory_2022] and adapting their behavior and physiology over time [@southwood_habitat_1977; @levins_evolution_1974]. Before or during periods of scarcity, organisms may cache resources [@nespolo_why_2022], build up fat reserves [@middleton_green-wave_2018], enter states of dormancy [@goldberg_hibernation_2021; @reher_short_2018; @mohr_cellular_2020], or even pause fetal growth [@schmidt_interplay_2020]. However, organisms may be unable to respond to changes in $\e(R)$ and $\var(R)$ optimally due to various reasons, including limited perceptive range [@steixner-kumar_strategies_2020], lack of experience [@foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @merkle_spatial_2019; @fagan_spatial_2013; @abrahms_memory_2019], avoidance of competitors and predators [@kacelnik_ideal_1992; @fretwell_territorial_1969], or a physiology that is not amenable to things like hibernation or fat storage. Thus, organisms may relocate their range to a sub-optimal location [@ciuti_effects_2012; @rich_anthropogenic_2012; @burson_competition_2018; @torrez-herrera_monkeys_2020], which may exacerbate the effects of $\e(R)$ and $\var(R)$ on both mean range size and the variance around it.

@@

# Conclusions

@@

# List of abbreviations

```{r abbreviations, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
library(dplyr)
read.csv('../writing/abbreviations.csv') %>%
  knitr::kable(format = 'latex', escape = FALSE) %>%
  kableExtra::row_spec(row = 0, bold = TRUE)
```

# Declarations

## Ethics approval and consent to participate

\noindent Not applicable.

## Consent for publication

\noindent Not applicable.

## Availability of data and materials

\noindent All code and data used for this manuscript is available on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch](https://github.com/QuantitativeEcologyLab/hr-resource-stoch), with the exception of two simulated datasets that were greater than 100 MB and the tapir data. The simulated data can be produced by running the scripts in the repository, while the tapir data is available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).

## Competing interests

\noindent The authors declare that they have no competing interests.

## Funding

\noindent SM was supported by funding from the University of British Columbia Okanagan, the Canadian Foundation for Innovation, BC Parks Living Labs, and MITACS. MJN was supported by NSERC Discovery Grant RGPIN-2021-02758 and the Canadian Foundation for Innovation. CHF was supported by NSF IIBR 1915347.

## Authors' contributions

\noindent SM performed the literature review, designed the simulations, analyzed the data, and wrote the manuscript. CHF contributed to the analyses. EPM provided the tapir telemetry data. MJN conceived the project idea and provided support throughout the analyses. All authors contributed to the writing and read and approved the final manuscript.

## Acknowledgements

\noindent We would like to thank Dr. Simon Wood for providing code to fit a Beta location-scale GAM despite not being involved directly with the project. Additionally, we thank all those who provided feedback on all posters, presentation, and writings related to this project. In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely, in alphabetical order by first name: Aimee Chhen, Jessa Marley, Kim Hinz, Lauren Mills, Sarah Wyse, and Dr. Simon Wood.

## Authors' information

\noindent See page 1 for affiliations.

\noindent Authors' emails:

\noindent SM: $\text{stefano.mezzini@ubc.ca}$

\noindent CHF: $\text{christen.fleming@ucf.edu}$

\noindent EPM: $\text{medici@ipe.org.br}$

\noindent MJN: $\text{michael.noonan@ubc.ca}$
